---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

options("ozgs.cache" = cachem::cache_disk("./readme_cache"))
```

# 📦 ozgs

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

`ozgs` fetches and caches `sf` geometries and data from the [Australian Statistical Geography Standard](https://www.abs.gov.au/statistics/standards/australian-statistical-geography-standard-asgs-edition-3/jul2021-jun2026/access-and-downloads/data-services-and-apis) (ASGS) web service.

## Installation

You can install the development version of `ozgs` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("gardiners/ozgs")
```

## Available geometries

Every geometry published by the ABS in every edition of the ASGS (1, 2 or 3) is currently available in `ozgs`.

These include the [Indigenous Structure](https://www.abs.gov.au/statistics/standards/australian-statistical-geography-standard-asgs-edition-3/jul2021-jun2026/indigenous-structure), [Statistical Areas](https://www.abs.gov.au/statistics/standards/australian-statistical-geography-standard-asgs-edition-3/jul2021-jun2026/main-structure-and-greater-capital-city-statistical-areas) (SA1, SA2, SA3 and SA4), [states and territories](https://www.abs.gov.au/statistics/standards/australian-statistical-geography-standard-asgs-edition-3/jul2021-jun2026/main-structure-and-greater-capital-city-statistical-areas/australia-and-stateterritory), [postcode approximations](https://www.abs.gov.au/statistics/standards/australian-statistical-geography-standard-asgs-edition-3/jul2021-jun2026/non-abs-structures/postal-areas) (POAs), [suburbs and localities](https://www.abs.gov.au/statistics/standards/australian-statistical-geography-standard-asgs-edition-3/jul2021-jun2026/non-abs-structures/suburbs-and-localities), [local government areas](https://www.abs.gov.au/statistics/standards/australian-statistical-geography-standard-asgs-edition-3/jul2021-jun2026/non-abs-structures/local-government-areas) (LGAs), [state](https://www.abs.gov.au/statistics/standards/australian-statistical-geography-standard-asgs-edition-3/jul2021-jun2026/non-abs-structures/state-electoral-divisions) and [federal](https://www.abs.gov.au/statistics/standards/australian-statistical-geography-standard-asgs-edition-3/jul2021-jun2026/non-abs-structures/commonwealth-electoral-divisions) electoral divisions, the [Remoteness Structure](https://www.abs.gov.au/statistics/standards/australian-statistical-geography-standard-asgs-edition-3/jul2021-jun2026/remoteness-structure) and the [mesh blocks](https://www.abs.gov.au/statistics/standards/australian-statistical-geography-standard-asgs-edition-3/jul2021-jun2026/main-structure-and-greater-capital-city-statistical-areas/mesh-blocks) from which each of these are constructed. 

`ozgs` uses [`arcgislayers`](https://github.com/R-ArcGIS/arcgislayers)📦 for easy querying, filtering and spatial queries.

## Caching

Requests are cached to avoid repeatedly fetching large quantities of data from the ABS servers. By default, `ozgs` creates a simple memory cache using `cachem::cache_mem()`, which persists for your current session. To persistently store ASGS geometries locally (or elsewhere), you can use any `cachem`-compatible cache function.

## Usage

Plot Tasmania's coastlines, using data from the 1st edition of the ASGS:

```{r example-tas}
library(ozgs)
library(ggplot2)

theme_set(theme_void())

tas <- get_geography("STE", edition = 1, where = "STE_NAME = 'Tasmania'")

ggplot(tas) +
  geom_sf()
```

Fetch every 3rd-edition SA1 that intersects the Sydney Local Government Area, at its extent in 2023:

```{r example-syd}
sydney_lga <- get_geography("LGA",
                            reference_date = 2023,
                            where = "LGA_NAME_2023 = 'Sydney'")

sydney_sa1 <- get_geography("SA1",
                            edition = 3,
                            filter_geom = sydney_lga$geometry)

ggplot(sydney_sa1) +
  geom_sf()
```

Store cached geometries within your working directory for reuse in other sessions (especially helpful if you're working in [RMarkdown](https://rmarkdown.rstudio.com/) or [Quarto](https://quarto.org/docs/computations/r.html), which start a fresh session on each render):

```{r}
project_cache <- cachem::cache_disk("./readme_cache")
aus <- get_geography("AUS", edition = 3, cache = project_cache)
```

Or, set `ozgs.cache` as an option for use everywhere in your script, with the same effect:

```{r}
options("ozgs.cache" = cachem::cache_disk("./readme_cache"))
```

